---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# washr

<!-- badges: start -->

[![R-CMD-check](https://github.com/openwashdata-dev/washr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/openwashdata-dev/washr/actions/workflows/R-CMD-check.yaml) [![Hippocratic License HL3-CORE](https://img.shields.io/static/v1?label=Hippocratic%20License&message=HL3-CORE&labelColor=5e2751&color=bc8c3d)](https://firstdonoharm.dev/version/3/0/core.html)

<!-- badges: end -->

The goal of `washr` is to provide tools to make the [openwashdata data package](https://openwashdata.org/pages/gallery/data/) workflow consistent and efficient.

## Installation

The easiest way to get started with `washr` is by installing it from CRAN:

```{r, eval=FALSE}
install.packages("washr")
```


To get the latest version, you can install the development version from [GitHub](https://github.com/) with:

``` {r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("openwashdata-dev/washr")
```

## How to use washr

To learn more about `washr`, check out the 'Get Started' page. This, along with the 'Reference' page, will give you a good overview of `washr`'s current capabilities and limitations.

If you decide to use `washr` for publishing your data (which we hope you will!), take a look at our detailed manual. It's regularly updated and walks you through the entire process step-by-step. The manual is your best bet for the most up-to-date information, as we're constantly improving it.
